Fail Faster

### 2025-03-18 paper/paper.tex
etna chart

  % in strict functionallanguages like OCaml and Scal
  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

@@

\subsection{Implementing and Testing Generators}
\begin{figure}[h]
  \includegraphics[scale=0.34]{allocs.png}\includegraphics[scale=0.334]{time.png}
\end{figure}
\label{subsection:pbt-for-pbt}
To establish the equivalence of generators up to program equality, 
we employ a PBT harness to compare their outputs over 1,000 random seeds. 
If a discrepancy arises, we report the unequal trees; if no counterexample is found,
we conclude that the generators exhibit equivalent behavior.
\subsection{Benchmarking speed \& resource usage}

\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints without bind (use map): lots of sampling, minimal binds.
    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
  \end{itemize}
}
\subsection{Impact on bug-finding ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques. 
Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.

Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
includes tasks that test insertion, deletion, and union operations on binary search trees).

We evaluate our case studies with both type-derived and custom generators.
Each generator has four implementations, which are equivalent up to program equality:
\begin {itemize}
  \item The original \bq generator (``BQ'').
  \item A staged version of the generator (``Staged'').
  \item A staged version of the generator using the C \rand (``Staged C'').
  \item A staged version incorporating the CSR \rand (``Staged CSR'').
\end{itemize}

To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.

\subsubsection{Variance between seeds}
The time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.

% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.

\subsubsection{Variance between trials}
Although the variance \textit{between} seeds in Etna can be very vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}

>>>>

  % in strict functionallanguages like OCaml and Scal
  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
optimize functional DSLs~\cn.
We apply this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

@@

\subsection{Implementing and Testing Generators}
\label{subsection:pbt-for-pbt}
To establish the equivalence of generators up to program equality, 
we employ a PBT harness to compare their outputs over 1,000 random seeds. 
If a discrepancy arises, we report the unequal trees; if no counterexample is found,
we conclude that the generators exhibit equivalent behavior.
\subsection{Benchmarking speed \& resource usage}
\begin{figure}[h]
  % \includegraphics[scale=0.33]{allocs.png}
  \includegraphics[scale=0.45]{time.png}
  \label{fig:benchmark-time}
  \caption{Average time taken to generate a value with the staged generator, in nanoseconds.}
\end{figure}
\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints without bind (use map): lots of sampling, minimal binds.
    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
  \end{itemize}
}
\subsection{Impact on bug-finding ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering. Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
includes tasks that test insertion, deletion, and union operations on binary search trees).

We evaluate our case studies with both type-derived and custom generators.
Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.

\subsubsection{Variance between seeds}
The time-to-failure of a given strategy on a given seed is not necessarily representative of 
the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
it is not perfect.
For example, it is theoretically possible
to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
if both generators fail to find the bug).
To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{}
\end{figure}
% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.

\subsubsection{Variance between trials}
Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
